---
title: "Predicting the MIP award winner"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Predicting the MIP award winner

**Problem**: The NBA is a sports league on the rise. It has a growing fanbase overseas (a professional league is currently being established in Africa with the help of the NBA), and since it lacks the stigma of CTE, it appears poised to steal some of the market share away from the NFL. There are a number of awards given out at the end of each season for various accomplishments, including the most improved player award (MIP). This award is given to the player who has shown the most progress during the regular season. Typically this means players that greatly increased their counting stats compared to their previous seasons in addition to overall player development. 

**Who would care about this?**: Sportwriters are the ones who vote for the award, and so are likely to be interested in seeing who the likely winners are, statistically speaking. Since there are so many different counting stats and a number of advanced statistics regarding playerse in the league, there are a number of fans who obsess over the numbers surrounding the game. Meaning sports analysts would be wise to base their speculation on statistics. Besides the media, sports betting agencies will want to be able to track a player's probability for winning the award in order to place the proper odds on the bet.

**Data Sources**: The data was pulled from a data-set on Kaggle, which in turned was scraped from *Basketball-reference* (https://www.kaggle.com/drgilermo/nba-players-stats/version/2). This data set contains player statistics from the past 67 seasons with 53 attributes, as well as some player information like the college team they played for and their home town. Another table was created from data for the award candidates, which was gathered from *Basketball-reference*. 

**Problem Solution**: In order to predict the award winner, building a logistic regression algorithm is likely the best approach. This algorithm will classify each input as either a MIP award winner or Not a MIP award winner, and it will give the probability for both classes. 

## Data Wrangling/Cleanup

The code for data wrangling and the rest of the prediction model can be found in the appendix. Data wrangling was performed using the tidyr and dplyr packages. New variables (MPG and PPG) were added to the player statistics data frame, and all player data prior to 1985 was removed.  The award was initiated the year before so all previous seasons did not have a MIP candidate, which could effect the predictive model. Also, a number of stats like turnovers and three point shots (the three pointer was even introduced until 79-80) were not tracked until the 1979-80 season. The two tables (player statistics and MIP candidates) were altered to match one another and then joined. NA values in the newly created MIP candidate column were converted to non-MIP. Blank columns were removed from the newly formed table. After reviewing MIP candidate mins and maxes, it was noticed that Dan Dickau had three MIP candidate seasons, all in 2005. The two rows with incomplete season stats were removed. New variables displaying the differences in all player statistics for the previous year were created. Here is a sample of the code used for that purpose

> MIP <- MIP %>% arrange(Player, Year) %>% mutate(dif_PPG = ifelse(Player == lag(Player), PPG - lag(PPG), 0))

Following the creation of these new variables, NA values for players in the year 1985 were replaced with zeros. Here is a summary of the cleaned up table

```{r MIP, echo = FALSE}
load(file = "MIP.RData")
summary(MIP)
```

## Statistical analysis and data visualization

Before visualizing the difference between MIP and non-MIP candidates in plots, the means for some of the statistics were viewed in a simple table, like so:

```{r table, echo=FALSE}
library(dplyr)
library(tidyr)
MIP %>% group_by(MIP_Candidate) %>% summarize(meandif_PPG = mean(dif_PPG))
```

Players who played less than 23 minutes per game were initially removed to better visualize the difference (as this was the minumum a MIP candidate played), but this data was kept for the regression model. 

Below is a simple line plot to depict the positive correlation between PPG and MPG. This is essentially true for all statistics. More playing time means more chances to score/steal/etc. It also typically means the player is better as well. This was before the minute parameter was implemented.

```{r plot, echo=FALSE}
library(ggplot2)
ggplot(MIP, aes(x = MPG, y = PPG)) + geom_line()
```

Here is a histogram depicting the frequency of PPG seasons (50 bins).

```{r plot2, echo=FALSE}
ggplot(MIP, aes(x = PPG)) + geom_histogram(bins = 50, aes(color = MIP_Candidate))
MIPplay <- MIP %>% subset(G > 55)
MIPplay <- MIP %>% subset(MPG > 23)
```

You can see that the average for MIP candidates is higher than the average for Non-candidate players, although it is hard to visualize. To make the comparison between MIP and non-MIP candidates clearer, players with fewer than 23 MPG and 55 G (games played) and players before 1985 were removed. 23 MPG was the minimum amount played for a MIP candidate, and 1985 was chosen because it was one year before the introduction of the award. 55 games played was chosen to remove players with limited data and extremely unlikely to be considered for the award. Below is the same plot but from the altered table.

```{r plot3, echo=FALSE}
ggplot(MIPplay, aes(x = PPG)) + geom_histogram(bins = 50, aes(color = MIP_Candidate))
```

Below is a scatter plot graphing PPG vs MPG for MIP and non-MIP candidates. There doesn’t appear to be a discernible trend for MIP vs non-MIP (although this is partly due to overcrowding). 

```{r plot4, echo=FALSE}
ggplot(MIPplay, aes(x = MPG, y = PPG)) + geom_point(aes(color = MIP_Candidate))
```

And here is a jittered scatter plot showing the difference in PPG with a color gradient depicting MPG. This does a slightly better job of depicting the difference between the two types. But a more obvious trend can be seen when looking at the difference in a players statistics from the previous season.

```{r plot5, echo=FALSE}
ggplot(MIPplay, aes(x = as.factor(MIP_Candidate), y = PPG)) + geom_jitter(aes(color = MPG)) + xlab("MIP candidacy")
```

The plot below graphs the difference in PPG vs the difference in MPG. The overall trend is very pronounced here as well as the differences between the two groups.

```{r plot6, echo=FALSE}
ggplot(MIPplay, aes(x = dif_MPG, y = dif_PPG)) + geom_point(aes(color = MIP_Candidate)) + xlab("Difference in MPG") + ylab("Difference in PPG")
```

Out of all the newly created variables, difference in VORP and difference in PPG appear the most significant.

```{r plot7, echo=FALSE}
ggplot(MIPplay, aes(x = dif_VORP, y = dif_PPG)) + geom_point(aes(color = MIP_Candidate)) + xlab("Difference in VORP") + ylab("Difference in PPG")
```

##Logistic Regression Model

Now that the data has been cleaned up and properly visualized, it is time to create the predictive model. To begin, the data was split (with a RNG for reproducability) into a training and a test set. 75% went to the training set and the other 25% to the testing set. The dependent variable, MIP Candidate, was converted to a 1 for "MIP candidate" and a 0 for "Non-MIP candidate", which was then converted into a factor. Virtually every counting stat available, including Age and the Year were used for the independent variables. Variables were removed to evaluate the effect on the AIC, which measures relative fit and favors a model with a lower number of parameters. The original model scored better by this metric than the second model which used only variables  The model was then used to predict MIP candidates for the test data set. Based on the coefficients, the variable which had the greatest impact on the model's performance was dif_PPG, followed by dif_VORP, WS, and MPG. The model classified an MIP candidate correctly 43% of the time, with a threshold value of 0.5. Here is a confusion matrixof the model.

```{r plot8, echo=FALSE}
load(file = "testtable.RData")
fourfoldplot(testtable)
```

And here is a plot of the ROC curve

```{r plot9, echo=FALSE}
load(file = "ROC.RData")
plot(roccurve)
```

The AUC is very high at 97%. This may imply that the model is overfitting. The model is very good at classifying non_MIP candidates, and fairly accurate at classifying MIP candidates. The vast majority of players who have been in the NBA will never win an MIP award, so this makes sense. The player with the highest probability to be an MIP candidate was CJ Mccollum in the 2015-2016 season at 97%. This makes sense as his dif in PPG was incredibly high. Here is a snapshot of the players with the highest MIP candidacy probability.

```{r plot12, echo=FALSE}
load(file = "MIPsample.RData")
library(knitr)
kable(MIPsample)
```

Many of these players were not only candidates, but actually won the award that year. The model seems to heavily favor dif_PPG, along with MPG (likely to differenctiate between bench/role players and starter caliber players). It also relies on a few advanced statistics like WS and VORP. 
##Random Forest Model
Another model was built using the random forest algorithm. Random Forest was chosen due to it's relative simplicity and because it is less likely to suffer from overfitting than the logistic regression model due to its use of randomization in the creation of subsets of the features applied to the model. To begin, a model was created with the same features as the logistic model. Here is a plot depicting feature importance.

```{r plot10, echo=FALSE}
library(randomForest)
load(file = "MIPforest.RData")
varImpPlot(MIPforest)
```

The mean decrease in accuracy measures how much the model's accuracy will decrease if the variable is removed. The mean decrease in gini measures the number of times a feature is used to split a node. This is also known as the mean decrease in impurity. Interestingly, the difference in PPG has less of an effect on the accuracy of the model than would be expected based on the results of the logistic regression model. However, dif_PPG obviously has an effect on the model based on the drastic reduction in impurity when it is included. The other features match the coefficients of the logit model. However, the predictive power of the random forest model is much less. Here is a confusion matrix of the model. 

```{r plot11, echo=FALSE}
load(file = "forestTable.RData")
fourfoldplot(forestTable)
```

But this is not the best way to compare the two models. The AUC for the random forest model was 98%, slightly higher than the regression model. Here is a plot of the ROC curve for random forest.

##Insert plot here


##Conclusions
Both models are able to pick an MIP candidate relatively accurately